Simulations

### 2018-07-12 1-preprocessing.Rmd
save

```{r}
file.copy(paste0("../paper1-packages/backingfiles/celiacQC", c(".bk", ".rds")),
          paste0("backingfiles/celiacQC", c(".bk", ".rds")))

library(bigsnpr)
str(celiac <- snp_attach("backingfiles/celiacQC.rds"), max.level = 1)
```

@@

```{r}
# Get population
pop.files <- list.files(path = "../thesis-celiac/Dubois2010_data/", 
                        pattern = "cluster_*", full.names = TRUE)
pop <- snp_getSampleInfos(celiac, pop.files)[[1]]
pop.names <- c("Netherlands", "Italy", "UK1", "UK2", "Finland")
celiac$fam$family.ID <- pop.names[pop]
celiac <- snp_save(celiac)
# Read modified file to check
celiac <- snp_attach("backingfiles/celiacQC.rds")
rle(celiac$fam$family.ID)
```

@@

```{r}
# Compute PCA
obj.svd <- snp_autoSVD(celiac$genotypes, 
                       celiac$map$chromosome,
                       celiac$map$physical.pos, 
                       ncores = nb_cores())

@@

```{r}
# New dataset with only non-outliers UK controls
length(ind.keep <- which(distM < THR))
celiac2 <- snp_attach(subset(celiac, ind.row = ind.keep))
obj.svd2 <- snp_autoSVD(celiac2$genotypes, 
                        celiac2$map$chromosome,
                        celiac2$map$physical.pos,
                        ncores = nb_cores())
plot(obj.svd2)
plot(obj.svd2, type = "scores") 
# Save this new PCA
saveRDS(obj.svd2, "backingfiles/PCA2.rds")
```

>>>>

```{r}
library(bigsnpr)
str(celiac <- snp_attach("backingfiles/celiacQC.rds"), max.level = 1)
```

@@

```{r}
# Get population
pop.files <- list.files(path = "backingfiles", 
                        pattern = "cluster_*", full.names = TRUE)
pop <- snp_getSampleInfos(celiac, pop.files)[[1]]
pop.names <- c("Netherlands", "Italy", "UK1", "UK2", "Finland")
celiac$fam$family.ID <- pop.names[pop]
celiac <- snp_save(celiac)
# Read modified file to check
celiac <- snp_attach("backingfiles/celiacQC.rds")
rle(celiac$fam$family.ID)
```

@@

```{r}
# Compute PCA
obj.svd <- snp_autoSVD(celiac$genotypes, 
                       celiac$map$chromosome,
                       celiac$map$physical.pos, 
                       ncores = nb_cores())
# Save this PCA
saveRDS(obj.svd, "backingfiles/PCA.rds")

@@

```{r}
# New dataset with only non-outliers UK controls
length(ind.keep <- which(distM < THR))
celiac2 <- snp_attach(subset(celiac, ind.row = ind.keep))
obj.svd2 <- snp_autoSVD(celiac2$genotypes, 
                        celiac2$map$chromosome,
                        celiac2$map$physical.pos,
                        ncores = nb_cores())
# Save this new PCA
saveRDS(obj.svd2, "backingfiles/PCA2.rds")

plot(obj.svd2)
plot(obj.svd2, type = "scores") 
```

### 2018-07-12 3-simus.Rmd
save

```{r, message=FALSE}
library(tidyverse)
library(glue)
library(bigsnpr)
# https://stackoverflow.com/q/9486952/6103040
options(bigstatsr.cluster.type = "PSOCK")
```

@@

    # Clumping on the test set
    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
                             ind.row = ind.test,
                             thr.r2 = 0.2, 
                             S = abs(gwas.train.gc$score),
                             size = 500,
                             is.size.in.bp = TRUE,
                             infos.pos = infos.pos,
                             ncores = NCORES)

    # PRS
    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
    lpS <- -predict(gwas.train.gc)
    nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
                   ind.test = ind.test,
                   ind.keep = ind.keep,
                   lpS.keep = lpS[ind.keep], 
                   thr.list = thrs)
  })[3]

@@

  lapply(1:3, function(i) {
    k <- indices[i]
    tibble(
      method = methods[i],
      eval = list(cbind(prs[, k], pheno.all[ind.test])),
      timing = timing,
      nb.preds = nb.pred[k]
    )
  }) %>%
    bind_rows()
}
```

@@

```{r}
logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test,
                       method = "logit-simple", alphas = 0.5) {

@@

    preds <- rowMeans(
      predict(cmsa.logit, X = G, ind.row = ind.test, 
                     covar.row = covar.all[ind.test, , drop = FALSE])
    )
  })[3]

  tibble(
    method   = method, 
    eval     = list(cbind(preds, pheno.all[ind.test])),
    timing   = timing,
    nb.preds = sum(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0)
  )
}
```

@@

```{r}
TTree <- "../../TTree-source/TTree"

@@

  tibble(
    method   = "T-Trees",
    eval     = list(preds[match(ind.test - 1, preds[[1]]), 2:3]),
    timing   = timing,
    nb.preds = sum(read.table(file.vim)$V2 != 0)
  )
}
```

### Logistic regression with biglasso

```{r}
logit.biglasso <- function(G3, pheno.all, covar.all, ind.train, ind.test) {

    biglasso <- biglasso(G3, pheno.all, ind.train, ncores = NCORES,
                         penalty = "enet", family = "binomial", alpha = 0.5)

    preds <- 1 / (1 + exp(-predict(biglasso, G3, ind.test)))
  })[3]

  aucs <- apply(preds, 2, bigstatsr::AUC, target = pheno.all[ind.test])
  ind.max <- which.max(aucs)

  tibble(
    method   = "biglasso", 
    eval     = list(cbind(preds[, ind.max], pheno.all[ind.test])),
    timing   = timing,
    nb.preds = sum(biglasso$beta[, ind.max] != 0)
  )
}
```

@@

  y.simu <- y.simu / sd(y.simu) * sqrt(h2)
  stopifnot(all.equal(drop(var(y.simu)), h2))
  y.simu <- y.simu + rnorm(nrow(G), sd = sqrt(1 - h2))
  pheno <- as.numeric(y.simu > qnorm(1 - K))
}
```

@@

  # Simulate phenotypes
  pheno.all <- get_pheno(
    G,    
    h2 = params[["par.h2"]], 
    M = as.integer(par.causal[1]), 
    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
    effects.dist = params[["par.dist"]], 
    model = params[["par.model"]], 
    K = 0.3                 
  )

  # Split in training/test sets
  ind.train <- sort(sample(n, size = params[["n.train"]]))
  ind.test <- setdiff(1:n, ind.train)

  # Get results from all methods
  res <- bind_rows(
    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
    logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple"),
    ttrees(TTree, "backingfiles/ttrees", pheno.all, ind.train, ind.test, n.trees = 100)
  )
  params[["res"]] <- list(res)
  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
}
```

@@

  # Simulate phenotypes
  pheno.all <- get_pheno(
    G,    
    h2 = params[["par.h2"]], 
    M = as.integer(par.causal[1]), 
    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
    effects.dist = params[["par.dist"]], 
    model = params[["par.model"]], 
    K = 0.3                 
  )

  # Split in training/test sets
  ind.train <- sort(sample(n, size = params[["n.train"]]))
  ind.test <- setdiff(1:n, ind.train)

@@

  # Simulate phenotypes
  pheno.all <- get_pheno(
    G6,    
    h2 = params[["par.h2"]], 
    M = as.integer(par.causal[1]), 
    ind.possible = `if`(par.causal[2] == "all", cols_along(G6), ind.HLA6),
    effects.dist = params[["par.dist"]], 
    model = params[["par.model"]], 
    K = 0.3                 
  )

  # Split in training/test sets
  ind.train <- sort(sample(n, size = params[["n.train"]]))
  ind.test <- setdiff(1:n, ind.train)

@@

### Scenario n°1 (comparison with biglasso)

```{r}
library(biglasso)
library(Matrix)
```

```{r, eval=FALSE}
G3 <- bigmemory::big.matrix(nrow(G), ncol(G) + ncol(covar.all),
                            backingfile = "G-PC",                 
                            backingpath = "backingfiles")
big_apply(G, function(X, ind) {
  G3[, ind] <- X[, ind]
  NULL
}, a.combine = 'c')
G3[, ncol(G) + cols_along(covar.all)] <- covar.all
```

```{r}
G3 <- bigmemory::attach.big.matrix("backingfiles/G-PC.desc")
```

```{r}
params.grid7 <- expand.grid(
  n.train = 6000,
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.8), 
  par.model  = c("simple"),
  num.simu   = 1:100,
  stringsAsFactors = FALSE
) 
```

```{r, eval=TRUE}
if (!dir.exists("results7")) dir.create("results7")

for (i in rows_along(params.grid7)) {

  res.file <- paste0("results7/simu_", i, ".rds")
  if (file.exists(res.file)) next

  params <- params.grid7[i, ]
  par.causal <- params[["par.causal"]][[1]]

  # Simulate phenotypes
  pheno.all <- get_pheno(
    G,    
    h2 = params[["par.h2"]], 
    M = as.integer(par.causal[1]), 
    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
    effects.dist = params[["par.dist"]], 
    model = params[["par.model"]], 
    K = 0.3                 
  )

  # Split in training/test sets
  ind.train <- sort(sample(n, size = params[["n.train"]]))
  ind.test <- setdiff(1:n, ind.train)

  # Get results from all methods
  res <- bind_rows(
    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test),
    logit.biglasso(G3, pheno.all, covar.all, ind.train, ind.test)
  )
  params[["res"]] <- list(res)
  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
}
```

### Scenario n°1 (comparison of different $\alpha$)

```{r, eval=TRUE}
if (!dir.exists("results8")) dir.create("results8")

for (i in rows_along(params.grid7)) {

  res.file <- paste0("results8/simu_", i, ".rds")
  if (file.exists(res.file)) next

  params <- params.grid7[i, ]
  par.causal <- params[["par.causal"]][[1]]

  # Simulate phenotypes
  pheno.all <- get_pheno(
    G,    
    h2 = params[["par.h2"]], 
    M = as.integer(par.causal[1]), 
    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
    effects.dist = params[["par.dist"]], 
    model = params[["par.model"]], 
    K = 0.3                 
  )

  # Split in training/test sets
  ind.train <- sort(sample(n, size = params[["n.train"]]))
  ind.test <- setdiff(1:n, ind.train)

  # Get results from all methods
  res <- bind_rows(
    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
               method = "logit-simple-enet-ridge", alphas = 0.001),
    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
               method = "logit-simple",            alphas = 0.5),
    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
               method = "logit-simple-lasso",      alphas = 1)
  )
  params[["res"]] <- list(res)
  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
}
```

>>>>

```{r, message=FALSE}
library(tidyverse)
library(glue)
library(bigsnpr)
```

@@

  })[3]

  lapply(c(0.05, 0.2, 0.8), function(thr.r2) {

    # Clumping
    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
                             ind.row = ind.test,
                             thr.r2 = thr.r2, 
                             S = abs(gwas.train.gc$score),
                             size = 500,
                             is.size.in.bp = TRUE,
                             infos.pos = infos.pos,
                             ncores = NCORES)

    # PRS
    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
    lpS <- -predict(gwas.train.gc)
    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
                   ind.test = ind.test,
                   ind.keep = ind.keep,
                   lpS.keep = lpS[ind.keep], 
                   thr.list = thrs)

@@

    lapply(1:3, function(i) {
      k <- indices[i]
      tibble(
        method = methods[i],
        pred = list(prs[, k]),
        thr.r2 = thr.r2,
        set = list(intersect(ind.keep, which(lpS > thrs[k]))),
        timing = timing
      )
    }) %>%
      bind_rows()
  }) %>%
    bind_rows()
}
```

@@

```{r}
logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method,
                       alphas = c(1, 0.5, 0.05, 0.001)) {

@@

    preds <- predict(cmsa.logit, X = G, ind.row = ind.test, 
                     covar.row = covar.all[ind.test, , drop = FALSE])
  })[3]

  tibble(
    method   = method, 
    pred     = list(preds),
    timing   = timing,
    alpha = attr(cmsa.logit, "alpha"),
    set = list(which(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0))
  )
}

```

@@

```{r}
bigstatsr:::assert_exist(TTree <- "../TTree-source/TTree")

@@

  tibble(
    method   = "T-Trees",
    pred     = list(preds[match(ind.test - 1, preds[[1]]), 2]),
    timing   = timing,
    set = list(which(read.table(file.vim)$V2 != 0))
  )
}
```

@@

  y.simu <- y.simu / sd(y.simu) * sqrt(h2)
  stopifnot(all.equal(drop(var(y.simu)), h2))
  y.simu <- y.simu + rnorm(nrow(G), sd = sqrt(1 - h2))
  list(pheno = as.integer(y.simu > qnorm(1 - K)), set = set, effects = effects)
}
```

@@

  # Simulate phenotypes
  simu_pheno <- get_pheno(
    G,    
    h2 = params[["par.h2"]], 
    M = as.integer(par.causal[1]), 
    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
    effects.dist = params[["par.dist"]], 
    model = params[["par.model"]]
  )
  pheno.all <- simu_pheno$pheno
  params[["set"]] <- list(simu_pheno$set)

  # Split in training/test sets
  ind.train <- sort(sample(n, size = params[["n.train"]]))
  ind.test <- setdiff(1:n, ind.train)
  params[["pheno"]] <- list(pheno.all[ind.test])

  # Get results from all methods
  res <- bind_rows(
    # PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
    logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple"),
    ttrees(TTree, "backingfiles/ttrees", pheno.all, ind.train, ind.test, n.trees = 100)
  )
  params[["res"]] <- list(res)
  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
}
```

@@

  # Simulate phenotypes
  simu_pheno <- get_pheno(
    G,    
    h2 = params[["par.h2"]], 
    M = as.integer(par.causal[1]), 
    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
    effects.dist = params[["par.dist"]], 
    model = params[["par.model"]]
  )
  pheno.all <- simu_pheno$pheno
  params[["set"]] <- list(simu_pheno$set)

  # Split in training/test sets
  ind.train <- sort(sample(n, size = params[["n.train"]]))
  ind.test <- setdiff(1:n, ind.train)
  params[["pheno"]] <- list(pheno.all[ind.test])

@@

  # Simulate phenotypes
  simu_pheno <- get_pheno(
    G,    
    h2 = params[["par.h2"]], 
    M = as.integer(par.causal[1]), 
    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
    effects.dist = params[["par.dist"]], 
    model = params[["par.model"]]
  )
  pheno.all <- simu_pheno$pheno
  params[["set"]] <- list(simu_pheno$set)

  # Split in training/test sets
  ind.train <- sort(sample(n, size = params[["n.train"]]))
  ind.test <- setdiff(1:n, ind.train)
  params[["pheno"]] <- list(pheno.all[ind.test])

@@
